Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-03-22 demystify.tex
updates

You should also talk about things that adcom members might not know about and can help make you different than the other 1000 applicants, e.g., you are member of a minority or LBGTQ+ group (\autoref{sec:urm}), your personal Github project with 1K+ stars or your regular contributions to well-known open-source projects (see \autoref{sec:improve-your-chance} for increasing your admission chance).

@@

\paragraph{Publications} A concrete evidence of research ability is having \emph{papers in reputable international journals or conferences}.
Having published papers, especially at top venues, is a sign that the applicant has successfully involved in research.  Students  accepted to top schools often have multiple first-authored papers at top venues.

@@

\begin{commentbox}[Vu:]
  Sometime PhD adcom in the US will share a document such as \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Foreign-Top-Schools}{this one}, which lists the top schools in several countries. In some cases we ask other faculty and students if we think they know about the place.  For example, when I was a postdoc at UMD, members in their CS PhD adcom ask me to evaluate applicants from Vietnam.  During my time at UNL and now here at GMU, I have looked at Vietnamese applications (whether they are assigned to me or not) and provide input to their reviewers, e.g., X is the top tech school in Vietnam and so it should be \emph{top} instead of \emph{unknown} foreign, which makes a huge difference.
\end{commentbox}
\begin{commentbox}[Deepak:]
  If an applicant is anxious about their school not being known outside their country, they can provide information about their school and department, with independent sources where such information can be verified.
\end{commentbox}

@@

Given the high number of quality applicants and limited number of spots, other than having a good application profile, you really want to show something that makes you \textbf{stand out}, e.g., do you have a degree or background in \emph{dance} or \emph{music} and want to integrate them with CS? are you a female or a minority in CS (research for ``URM minority in CS in the US'' on Google) (\autoref{sec:urm})? Do you participate in outreach activities that help increase diversity and inclusion in CS? All of these are unique and would get noticed from reviewers.

Even if you do not have research experience, you can talk about your personal projects, as long as they can help show you can do research. For example, if you have an open-source project on Github that is used by many people, has lots of stars in Github, do talk about it. If you often write technical, research-like blogs with many viewers, talk about them too.

In his \href{https://matt.might.net/articles/how-to-apply-and-get-in-to-graduate-school-in-science-mathematics-engineering-or-computer-science/}{post}, Matt Might was initially unsure about an application. However, upon learning that the applicant had led a 100km hike in the Himalayas, he decided to accept the applicant.  This is a good example of being \emph{stand out}, and I would also advocate for that student as this shows they have the persistence and determination required for research.

@@

\section{Member of an underrepresented or LBGTQ+ groups}\label{sec:urm}
A common question is whether you should mention that you are a member of an underrepresented and minority (URM) group\footnote{URM in CS in the US include African Americans, Hispanic Americans, American Indians/Native Alaskans, Native Hawaiians/Pacific Islanders (excluding Asian Americans), and multiracial students identifying with at least one of the previously listed URM categories.} or LBGTQ+ group (, e.g., in your SOP).
This is a strong concern for international (and even domestic) students, who may have encountered discrimination in their countries or personal experiences.
%This question arises often for international (and even domestic) students. The main worry is that they might get discriminated against, as often happen in their countries or personal experiences. %, and (ii) mentioning this might make them look like they are asking for special treatment.

In my personal opinion, highlighting URM or LBGTQ+ identity can \emph{indeed be beneficial}, especially if you can articulate how your diverse experiences contribute to diversity and inclusion in academic.  Many universities have a strong commitment to diversity and inclusion and actively
\emph{recruit students from underepresentative backgrounds} (e.g., some \href{https://cse-climate.engin.umich.edu/reports/climate-dei-reports/cse-climate-and-dei-report-2022-2023/#grad-ethnicity}{stats from UMich CSE} and \href{https://www.gmu.edu/news/2022-09/mason-now-top-10-public-university-diversity-innovation-and-cybersecurity-education-us}{GMU} often promotes itself as one of the most diverse universities in the US). Moreover, many scholarships and fellowships are created specifically for URM and LBGTQ+ students, which you should explore if qualify.

@@

\paragraph{Research Experiences \autoref{sec:undergrad-research}} You might have many opportunities to do research as an undergraduates, e.g., through REUs and internships at your undergrad university.  Highlight such experiences in your application.

@@

\chapter{Research Opportunities for Undergraduates}\label{sec:undergrad-research}

As discussed in \autoref{sec:undergrad-research}, research experience is a crucial component of a strong PhD application. It gives you opportunities to try out research, determine what research area you're interested in, publish papers (\autoref{sec:research-experiences}), and get strong LoRs (\autoref{sec:lor}).

 This section provides some guidance on how to gain research experience as an undergraduate student.  It also gives advice to international students or domestic ones from small colleges where research opportunities are limited.

@@

By contributing code, fixing bugs, implementing new features, or providing documentation, students can gain practical research experience and interact with experienced developers and researchers in the field. Not only you gain research experience, you might be able to get a LoR from the project maintainer and definitely should write about this experience in your SOP (\autoref{sec:research-statement}.)

\paragraph{Virtual Research Opportunities}: This is less common but several places offer virtual internships and research programs aimed at providing hands-on research experience. These programs often involve working remotely under the guidance of experienced mentors and collaborating with a team of fellow researchers. For instance, \href{https://docs.google.com/forms/d/1btIwt4HwjyKMOUk-EMy3rbkfWzFxv2lNrMm_zkd0pA4/viewform?edit_requested=true}{UIUC+ Summer Undergraduate Research in Software Engineering}  offers an unpaid remote internship for software engineering students all over the world to collaborate with mentors from University of Illinois at Urbana Champaign.

>>>>

You should also talk about things that adcom members might not know about and can help make you \textbf{stand out} in the application pool of  thousands of applicants, e.g., you are member of a minority or LGBTQ+ group (\autoref{sec:urm}), your personal Github project with 1K+ stars or your regular contributions to well-known open-source projects (see \autoref{sec:improve-your-chance} for increasing your admission chance).

@@

\paragraph{Publications} A concrete evidence of research ability is having \emph{papers in reputable international journals or conferences}.
Having published papers, especially at top venues, is a sign that the applicant has successfully involved in research.  Applicants  admitted to top schools often have multiple first-authored papers at top places.

@@

\begin{commentbox}[Vu:]
  Sometime PhD adcom in the US will share a document such as \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Foreign-Top-Schools}{this one}, which lists the top schools in several countries. We also ask other faculty and students if we think they know about the place.  For example, when I was a postdoc at UMD, members in their CS PhD adcom ask me to evaluate applicants from Vietnam.  During my time at UNL and now here at GMU, I have looked at Vietnamese applications (whether they are assigned to me or not) and provide input to their reviewers, e.g., X is the top tech school in Vietnam and so it should be \emph{top} instead of \emph{unknown} foreign, which makes a huge difference.
\end{commentbox}
\begin{commentbox}[Deepak:]
  If an applicant is anxious about their school not being known outside their country, they can provide information about their school and department, with independent sources where such information can be verified.
\end{commentbox}

@@

Given the high number of quality applicants and limited number of spots, in addition to having a good application profile, you really want to show something that makes you \textbf{stand out}.  For example, even if you do not have research experience, you can talk about your personal projects, as long as they can help show you can do research. For example, if you have an open-source project on Github that is used by many people, has lots of stars in Github, do talk about it. If you often write technical, research-like blogs with many viewers, talk about them too.

There are other things you might not think are important but can make you stand out. For example, if you have a strong background in a field that is not CS but can be integrated with CS, e.g., you have a degree or background in \emph{dance} or \emph{music} and want to integrate them with CS, do talk about it. Are you a female or a minority in CS (\autoref{sec:urm})? Do you participate in outreach activities that help increase diversity and inclusion in CS? Diversity is highly valued in CS programs in the US, and experiences in these areas can make you stand out and get noticed from reviewers.

\begin{commentbox}[Vu:]
In his \href{https://matt.might.net/articles/how-to-apply-and-get-in-to-graduate-school-in-science-mathematics-engineering-or-computer-science/}{post}, Matt Might was initially unsure about an application. However, upon learning that the applicant had led a 100km hike in the Himalayas, he decided to accept the applicant.  This is a good example of being \emph{stand out}, and I would also advocate for that student as this shows they have the persistence and determination required for research.
\end{commentbox}

@@

\section{Member of an underrepresented or LGBTQ+ groups}\label{sec:urm}
A common question is whether you should mention that you are a member of an underrepresented and minority (URM) group\footnote{URM in CS in the US include African Americans, Hispanic Americans, American Indians/Native Alaskans, Native Hawaiians/Pacific Islanders (excluding Asian Americans), and multiracial students identifying with at least one of the previously listed URM categories.} or LGBTQ+ group (, e.g., in your SOP).
This is a strong concern for international (and even domestic) students, who may have encountered discrimination in their countries or personal experiences.
%This question arises often for international (and even domestic) students. The main worry is that they might get discriminated against, as often happen in their countries or personal experiences. %, and (ii) mentioning this might make them look like they are asking for special treatment.

In my personal opinion, highlighting URM or LGBTQ+ identity can \emph{indeed be beneficial}, especially if you can articulate how your diverse experiences contribute to diversity and inclusion in academic.  Many universities have a strong commitment to diversity and inclusion and actively
\emph{recruit students from underepresentative backgrounds} (e.g., some \href{https://cse-climate.engin.umich.edu/reports/climate-dei-reports/cse-climate-and-dei-report-2022-2023/#grad-ethnicity}{stats from UMich CSE} and \href{https://www.gmu.edu/news/2022-09/mason-now-top-10-public-university-diversity-innovation-and-cybersecurity-education-us}{GMU} often promotes itself as one of the most diverse universities in the US). Moreover, many scholarships and fellowships are created specifically for URM and LGBTQ+ students, which you should explore if qualify.

@@

\paragraph{Research Experiences \autoref{sec:research-opportunities}} You might have many opportunities to do research as an undergraduates, e.g., through REUs and internships at your undergrad university.  Highlight such experiences in your application.

@@

\chapter{Research Opportunities}\label{sec:research-opportunities}

As discussed in \autoref{sec:research-opportunities}, research experience is a crucial component of a strong PhD application. It gives you opportunities to try out research, determine what research area you're interested in, publish papers (\autoref{sec:research-experiences}), connect with researchers and get strong LoRs (\autoref{sec:lor}).

 This section provides some guidance on how to gain research experience, e.g., as an undergraduate student.  It also gives advice to students from small colleges where research opportunities are limited.

@@

By contributing code, fixing bugs, implementing new features, or providing documentation, students can gain practical research experience and interact with experienced developers and researchers in the field. Not only you gain research experience, you might be able to get a LoR from the project maintainer, and you should write about this experience in your SOP (\autoref{sec:research-statement}.)

\paragraph{Virtual Research Opportunities} This is less common but several places offer virtual internships and research programs aimed at providing hands-on research experience. These programs often involve working remotely under the guidance of experienced mentors and collaborating with a team of fellow researchers. For instance, \href{https://docs.google.com/forms/d/1btIwt4HwjyKMOUk-EMy3rbkfWzFxv2lNrMm_zkd0pA4/viewform?edit_requested=true}{UIUC+ Summer Undergraduate Research in Software Engineering}  offers an unpaid remote internship for software engineering students all over the world to collaborate with mentors from University of Illinois at Urbana Champaign.
